Indexing Variation Graphs\footnote{This work was supported by the Wellcome Trust grant [098051].}

### 2016-04-12 paper/paper.tex
Rewrote parts of the existing text.

\begin{abstract}% FIXME
Lorem ipsum dolor sit amet, consectetur adipiscing elit. Praesent convallis orci arcu, eu mollis dolor. Aliquam eleifend suscipit lacinia. Maecenas quam mi, porta ut lacinia sed, convallis ac dui. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Suspendisse potenti.
 \end{abstract}

@@

Manifold DBG: \cite{Lin2014}

@@

In some cases, we also consider potentially \emph{infinite} character sequences $S = (s_{i})_{i \in Z}$, where set $Z$ is a contiguous subset of $\mathbb{Z}$. The notion of substrings generalizes for such infinite sequences in a natural way. We say that a substring of an infinite sequence $S$ is \emph{left-infinite} if it extends infinitely to the left, and \emph{right-infinite} if it extends infinitely to the right. A substring of a finite or infinite sequence $S$ is \emph{left-maximal} if it is left-infinite or a prefix; \emph{right-maximal} if it is right-infinite or a suffix; and \emph{maximal} if it is both left-maximal and right-maximal.

@@

A \emph{graph} $G = (V, E)$ consists of a set of \emph{nodes} $V = \set{0, \dotsc, \abs{V}-1}$ and a set of \emph{edges} $E \subseteq V \times V$. We say that $(u, v) \in E$ is an edge \emph{from} node $u$ \emph{to} node $v$, and assume that the edges are \emph{directed}: $(u, v) \ne (v, u)$ for $u \ne v$. The \emph{indegree} $G.\gindegree(v)$ of node $v$ is the number of \emph{incoming} edges to node $v$, while the \emph{outdegree} $G.\goutdegree(v)$ is the number of \emph{outgoing} edges from node $v$.

The graphs we consider are \emph{labeled} with alphabet $\Sigma$: each node $v \in V$ has a \emph{label} $G.\glabel(v) \in \Sigma$. A \emph{path} in a graph is a sequence of nodes $P = v_{0} \dotsm v_{\abs{P}-1}$ such that $(v_{i}, v_{i+1}) \in E$ for all $i, i+1 \in \set{0, \dotsc, \abs{P}-1}$. The label of a path is the concatenation of node labels $G.\glabel(P) = G.\glabel(v_{0}) \dotsm G.\glabel(v_{\abs{P}-1})$. We can also generalize the definition for infinite paths $P = (v_{i})_{i \in Z}$ as we did with infinite character sequences in Section~\ref{sect:strings}.

\begin{definition}[\kcollection{k}]
Let $\mathcal{S}$ be a collection of character sequences over alphabet $\Sigma$, and let $k$ be a parameter value. We say that $\mathcal{S}$ is a \emph{\kcollection{k}}, if each sequence $S \in \mathcal{S}$
\begin{enumerate}
\item is left-infinite or begins with $\#^{k}$;
\item is right-infinite or ends with $\$^{k}$; and
\item contains no other occurrences of characters $\#$ and $\$$,
\end{enumerate}
and the collection contains substrings $\#^{k}$ and $\$^{k}$.
\end{definition}

% FIXME example of strings, DBG, long paths, false positives
% FIXME some comments about DBGs
\begin{definition}[de~Bruijn graph]
Let $\mathcal{S}$ be a \kcollection{k} over alphabet $\Sigma$. The \orderk{k} \emph{de~Bruijn graph} of $\mathcal{S}$ is a graph $G = (V, E)$ such that
\begin{enumerate}
\item each node $v_{X} \in V$ represent a distinct \kmer{k} occurring in $\mathcal{S}$;
\item $G.\glabel(v_{X}) = X[0]$ and $G.\gkey(v_{X}) = X$ for all $v_{X} \in V$; and
\item each edge $(v_{X}, v_{X'}) \in E$ represents a distinct \kmer{k+1} $Y = Xc = c'X'$ (where $c, c' \in \Sigma$) occurring in $\mathcal{S}$.
\end{enumerate}
We use the node corresponding to $\#^{k}$ as the source node $s$ and the node corresponding to $\$^{k}$ as the sink node $t$, adding the technical edge $(t, s)$ in the usual way.
\end{definition}

\begin{proof}
Let $X = S[i, i+\abs{X}-1]$ for a sequence $S \in \mathcal{S}$. For $i \le j \le i+\abs{X}-1$, there is a node $v_{j} \in V$ representing substring $X_{j} = S[j, j+k-1]$, with $G.\glabel(v_{j}) = S[j]$. By definition, the de~Bruijn graph has an edge $(v_{j}, v_{j+1})$ for $i \le j < i+\abs{X}-1$. Hence $P = v_{i} \dotsm v_{i+\abs{X}-1}$ is a path in $G$ and $G.\glabel(P) = X$.
\end{proof}

\begin{lemma}[No false positives with $\abs{P} \le k+1$]\label{lemma:dbg-fp}
Let $G = (V, E)$ be an \orderk{k} de~Bruijn graph of \kcollection{k} $\mathcal{S}$, and let $P$ be a path in $G$ with $\abs{P} \le k+1$. Then the label of the path is a substring of a sequence $S \in \mathcal{S}$.
\end{lemma}

\begin{proof}
If $\abs{P} = 0$, the label is an empty string and hence a substring of each sequence $S \in \mathcal{S}$.
If $\abs{P} = 1$, the label is equal to $G.\glabel(v_{0})$, which is a prefix of $G.\gkey(v_{0})$ and hence a substring of $\mathcal{S}$.
Now assume that $G.\glabel(P) = cX$, where $c \in \Sigma$ and $1 \le \abs{X} \le k$. By induction, string $X$ is a prefix of $G.\gkey(v_{1})$. As edge $(v_{0}, v_{1})$ exists, $c \cdot G.\gkey(v_{1})$ must be a substring of $\mathcal{S}$. Hence $cX$ is a substring of $\mathcal{S}$, and also a prefix of $G.\gkey(v_{0})$ if $\abs{cX} \le k$.
\end{proof}

@@

The \emph{Burrows-Wheeler transform (BWT)} \cite{Burrows1994} is a permutation of the text sharing the combinatorial structure of the suffix array. The BWT of text $T[0, n-1]$ is a string $\BWT[0, n-1]$ defined as $\BWT[i] = T[(\SA[i]-1) \bmod n]$. Given the \emph{lexicographic rank} $i$ of suffix $T[\SA[i], n-1]$, we can use \emph{\LFmapping} on the BWT to find the lexicographic rank of the previous suffix $T[(\SA[i]-1) \bmod n, n-1]$. Let
$$
\LF(i) = \Carray[\BWT[i]] + \BWT.\rank(i, \BWT[i]),
$$
where $\Carray[c]$ is the number of suffixes of the text starting with any character $c' < c$. Then $\SA[\LF(i)] = (\SA[i]-1) \bmod n$. We can also generalize the definition for any character $c \in \Sigma$:
$$
\LF(i, c) = \Carray[c] + \BWT.\rank(i, c).
$$
Let $X$ be a string. If there are $i$ suffixes $S'$ of text $T$ such that $S' < X$ in lexicographic order, then there are $\LF(i, c)$ suffixes smaller than string $cX$ in lexicographic order.

@@

@@

A \emph{path index} is a generalization of text indexes for \emph{labeled graphs}. Given a path index for graph $G = (V, E)$, we want to use the index to find the \emph{start nodes} $v_{0} \in V$ of the paths $P = v_{0} \dotsm v_{\abs{X}-1}$ \emph{matching} the pattern $X$ (paths $P$ with $G.\glabel(P) = X$).

@@

We can build a \kcollection{k} of maximal path labels by starting with the collection $\mathcal{S}$ of maximal path labels. In order to transform $\mathcal{S}$ into a \kcollection{k}, we adjust the number of characters $\#$ at the beginning of each non-left-infinite sequence, and the number of characters $\$$ at the end of each non-right-infinite sequence. We also add string $\#^{k} \$^{k}$ into the collection if necessary.

\begin{definition}[de~Bruijn graph of a graph]
Let $G = (V, E)$ be a labeled graph, and let $\mathcal{S}$ be a \kcollection{k} of maximal path labels in $G$. The \orderk{k} de~Bruijn graph of $\mathcal{S}$ is also the \orderk{k} de~Bruijn graph of graph $G$.
\end{definition}

@@

On the other hand, we can use the \orderk{k} de~Bruijn graph $G' = (V', E')$ of graph $G = (V, E)$ directly as a \kmer{k} index of $G$. For each key-value pair $(X, W)$ in the index, we find the node $v_{X} \in V'$ with $G'.\gkey(v_{X}) = X$, and attach the \emph{value} $G'.\gvalue(v_{X}) = W$ to the node. We can then search for patterns of any length in the de~Bruijn graph. As Lemmas~\ref{lemma:dbg-fn} and \ref{lemma:dbg-fp} suggest, there will be no false negatives, and no \emph{false positives} with short enough patterns. With longer patterns, we have to \emph{verify} the results in graph $G$ to avoid false positives. We will develop the idea of using graphs similar to de~Bruijn graphs as path indexes further in the next section.

@@

\begin{definition}[Path graph]
Let $\mathcal{S}$ be a \kcollection{k} of the labels of maximal paths in graph $G = (V, E)$, and let $\mathcal{K}$ be a prefix-free set of strings containing $\#^{k}$ and $\$^{k}$. We assume that each right-maximal substring $S$ in $\mathcal{S}$ prefix-matches a string $K \in \mathcal{K}$, and that each string $K \in \mathcal{K}$ is a substring of $\mathcal{S}$. Then the \orderk{k} \emph{path graph} of graph $G$ with \emph{key set} $\mathcal{K}$ is a graph $G' = (V', E')$ such that
\begin{enumerate}
\item each node $v_{K} \in V'$ represent a distinct key $K \in \mathcal{K}$;
\item $G'.\glabel(v_{K}) = K[0]$ and $G'.\gkey(v_{K}) = K$ for all $v_{K} \in V$; and
\item each edge $(v_{K}, v_{K'}) \in E'$ represents the occurrence of substring $K[0] K'$ in $\mathcal{S}$ such that strings $K$ and $K[0] K'$ prefix-match.
\end{enumerate}
We use the node corresponding to $\#^{k}$ as the source node $s$ and the node corresponding to $\$^{k}$ as the sink node $t$, adding the technical edge $(t, s)$ in the usual way.
\end{definition}

% FIXME from here
- k-index: has a node matching any k-path in the input graph
- edge condition: Y has predecessor with c -> edge (X, Y) if cY matches X
- k-equivalent: union of values for nodes matching X identical for all |X| <= k
- k-equivalent + edge condition: no false positives of length <= k; no false negatives

pruned de Bruijn graphs: merge nodes of DBG while keeping the path graph k-equivalent

using pruned de Bruijn graphs as indexes; false positives
- string Y matches node with key X: X is a prefix of Y or Y is a prefix of X
- valid for existence queries of length up to k+1 and locate queries of length up to k

@@

I want to thank \dots

>>>>

\begin{abstract}
Variation graphs, which represent genetic variation within a population, are replacing sequences as reference genomes. Path indexes are one of the most important tools for working with variation graphs. They generalize text indexes for graphs, allowing one to find the paths matching the query string. We propose using pruned de Bruijn graphs as path indexes, encoding them space-efficiently with the Burrows-Wheeler transform. We also generalize many ideas from text indexing literature to work with graphs. The proposed approach has been implemented in the variation graph toolkit vg.
\end{abstract}

@@

@@

In some cases, we also consider potentially \emph{infinite} character sequences $S = (s_{i})_{i \in Z}$, where set $Z$ is a contiguous infinite subset of $\mathbb{Z}$. The notion of substrings generalizes for such infinite sequences in a natural way. We say that a substring of an infinite sequence $S$ is \emph{left-infinite} if it extends infinitely to the left, and \emph{right-infinite} if it extends infinitely to the right. A substring of a finite or infinite sequence $S$ is \emph{left-maximal} if it is left-infinite or a prefix; \emph{right-maximal} if it is right-infinite or a suffix; and \emph{maximal} if it is both left-maximal and right-maximal.

@@

The following definitions are useful when discussing text indexes and path indexes.

@@

The \emph{Burrows-Wheeler transform (BWT)} \cite{Burrows1994} is a permutation of the text with the same combinatorial structure as the suffix array. Given text $T[0, n-1]$, its Burrows-Wheeler transform is a string $\BWT[0, n-1]$, where $\BWT[i] = T[(\SA[i]-1) \bmod n]$. Given the \emph{lexicographic rank} $i$ of suffix $T[\SA[i], n-1]$, we can use \emph{\LFmapping} on the BWT to find the lexicographic rank of the previous suffix $T[(\SA[i]-1) \bmod n, n-1]$. Let
$$
\LF(i) = \Carray[\BWT[i]] + \BWT.\rank(i, \BWT[i]),
$$
where $\Carray[c]$ is the number of suffixes of the text starting with any character $c' < c$. Then $\SA[\LF(i)] = (\SA[i]-1) \bmod n$. We can also generalize the definition for any character $c \in \Sigma$:
$$
\LF(i, c) = \Carray[c] + \BWT.\rank(i, c).
$$
Let $X$ be a string. If there are $i$ suffixes $S'$ of text $T$ such that $S' < X$ in lexicographic order, then there are $\LF(i, c)$ suffixes smaller than string $cX$ in lexicographic order.

@@

A \emph{graph} $G = (V, E)$ consists of a set of \emph{nodes} $V = \set{0, \dotsc, \abs{V}-1}$ and a set of \emph{edges} $E \subseteq V \times V$. We say that $(u, v) \in E$ is an edge \emph{from} node $u$ \emph{to} node $v$, and assume that the edges are \emph{directed}: $(u, v) \ne (v, u)$ for $u \ne v$. The \emph{indegree} $G.\gindegree(v)$ of node $v$ is the number of \emph{incoming} edges to $v$, while the \emph{outdegree} $G.\goutdegree(v)$ is the number of \emph{outgoing} edges from $v$.

The graphs we consider are \emph{labeled} with alphabet $\Sigma$: each node $v \in V$ has a \emph{label} $G.\glabel(v) \in \Sigma$. A \emph{path} in a graph is a sequence of nodes $P = v_{0} \dotsm v_{\abs{P}-1}$ such that $(v_{i}, v_{i+1}) \in E$ for all $i, i+1 \in \set{0, \dotsc, \abs{P}-1}$. We say that $v_{0}$ is the \emph{start} node and $v_{\abs{P}-1}$ is the \emph{end} node of the path. The label of a path is the concatenation of node labels $G.\glabel(P) = G.\glabel(v_{0}) \dotsm G.\glabel(v_{\abs{P}-1})$. We can also generalize the definition for infinite paths $P = (v_{i})_{i \in Z}$ as we did with infinite character sequences in Section~\ref{sect:strings}.

In this paper, we work with de~Bruijn graphs and their generalizations. In order to do so, we need to define collections of (finite or infinite) sequences suitable for constructing \orderk{k} de~Bruijn graphs.

\begin{definition}[\kcollection{k}]
Let $\mathcal{S}$ be a collection of character sequences over alphabet $\Sigma$, and let $k$ be a parameter value. We say that $\mathcal{S}$ is a \emph{\kcollection{k}}, if the collection contains substrings $\#^{k}$ and $\$^{k}$, and each sequence $S \in \mathcal{S}$
\begin{enumerate}
\item is left-infinite or begins with $\#^{k}$;
\item is right-infinite or ends with $\$^{k}$; and
\item contains no other occurrences of characters $\#$ and $\$$.
\end{enumerate}
\end{definition}

% FIXME example of strings, DBG, long paths, false positives
\begin{definition}[de~Bruijn graph]
Let $\mathcal{S}$ be a \kcollection{k} over alphabet $\Sigma$. The \orderk{k} \emph{de~Bruijn graph} of $\mathcal{S}$ is a graph $G = (V, E)$ such that
\begin{enumerate}
\item each node $v_{X} \in V$ represent a distinct \kmer{k} $X$ occurring in $\mathcal{S}$, with $G.\glabel(v_{X}) = X[0]$;
\item each node $v_{X} \in V$ has a \emph{key} $G.\gkey(v_{X}) = X$; and
\item each edge $(v_{X}, v_{Y}) \in E$ represents a \kmer{k+1} $X[0]Y = Xc$ (where $c \in \Sigma$) occurring in $\mathcal{S}$.
\end{enumerate}
We use the node corresponding to $\#^{k}$ as the source node $s$ and the node corresponding to $\$^{k}$ as the sink node $t$, adding the technical edge $(t, s)$ in the usual way.
\end{definition}

The following lemmas summarize the key properties of de~Bruijn graphs for indexing purposes.

\begin{lemma}[Keys prefix-match with path labels]\label{lemma:dbg-key}
Let $G = (V, E)$ be an \orderk{k} de~Bruijn graph, and let $P$ be a path starting from node $v_{0} \in V$. Then strings $G.\gkey(v_{0})$ and $G.\glabel(P)$ prefix-match.
\end{lemma}

\begin{proof}
Let $P[0, k'-1] = v_{0} \dotsm v_{k'-1}$ be a prefix of path $P$, where $1 \le k' \le \min(k, \abs{P})$.
If $k' = 1$, we know that $G.\glabel(P[0]) = G.\glabel(v_{0})$ is a prefix of $G.\gkey(v_{0})$.
Now assume that $k' > 1$. By induction, we know that $G.\glabel(P[1, k'-1])$ is a prefix of $G.\gkey(v_{1})$. As edge $(v_{i}, v_{i+1}) \in E$ exists, $G.\glabel(v_{0}) \cdot G.\gkey(v_{1}) = G.\gkey(v_{0}) \cdot c$ (where $c \in \Sigma$), and hence $G.\glabel(P[0, k'-1])$ is a prefix of $G.\gkey(v_{0})$. As the first $\min(k, \abs{P})$ characters of $G.\gkey(v_{0})$ and $G.\glabel(P)$ are identical, the result follows.
\end{proof}

\begin{proof}
Let $X = S[i, i+\abs{X}-1]$ for a sequence $S \in \mathcal{S}$. For $i \le j \le i+\abs{X}-1$, there is a node $v_{j} \in V$ representing substring $S[j, j+k-1]$, with $G.\glabel(v_{j}) = S[j]$. By definition, the de~Bruijn graph has an edge $(v_{j}, v_{j+1})$ for $i \le j < i+\abs{X}-1$. Hence $P = v_{i} \dotsm v_{i+\abs{X}-1}$ is a path in $G$ and $G.\glabel(P) = X$.
\end{proof}

\begin{lemma}[No false positives with $\abs{P} \le k$]\label{lemma:dbg-fp}
Let $G = (V, E)$ be an \orderk{k} de~Bruijn graph of \kcollection{k} $\mathcal{S}$, and let $P = v_{0} \dotsm v_{\abs{P}-1}$ be a path in $G$ with $\abs{P} \le k$. Then the label of the path is a prefix of $G.\gkey(v_{0})$ and hence a substring of a sequence $S \in \mathcal{S}$.
\end{lemma}

\begin{proof}
The label $G.\glabel(P)$ prefix-matches $G.\gkey(v_{0})$ by Lemma~\ref{lemma:dbg-key}. As $\abs{P} \le k$, string $G.\glabel(P)$ is a prefix of $G.\gkey(v_{0})$
\end{proof}

Note that the labels of all paths of length $k+1$ are also substrings of collection $\mathcal{S}$.

@@

A \emph{path index} is a generalization of text indexes for \emph{labeled graphs}. Given a path index for graph $G = (V, E)$, we want to use the index to find the start nodes $v_{0} \in V$ of the paths $P = v_{0} \dotsm v_{\abs{X}-1}$ \emph{matching} the pattern $X$ (paths $P$ with $G.\glabel(P) = X$).

@@

We can build a \kcollection{k} based on the labels of maximal paths by starting with the collection $\mathcal{S}$ of such labels. In order to transform $\mathcal{S}$ into a \kcollection{k}, we adjust the number of characters $\#$ at the beginning of each non-left-infinite sequence, and the number of characters $\$$ at the end of each non-right-infinite sequence.

\begin{definition}[de~Bruijn graph of a graph]
Let $G = (V, E)$ be a labeled graph, and let $\mathcal{S}$ be a \kcollection{k} of maximal path labels in $G$. Then the \orderk{k} de~Bruijn graph of $\mathcal{S}$ is the \orderk{k} de~Bruijn graph of graph $G$.
\end{definition}

@@

On the other hand, we can use the \orderk{k} de~Bruijn graph $G' = (V', E')$ of graph $G = (V, E)$ as a \kmer{k} index of $G$ that supports queries of arbitrary length. For each key-value pair $(X, W)$ in the index, we find the node $v_{X} \in V'$ with $G'.\gkey(v_{X}) = X$, and attach the \emph{value} $G'.\gvalue(v_{X}) = W$ to the node.

\begin{definition}[de~Bruijn graph as an index]
Let $G = (V, E)$ be a graph, and let $G' = (V', E')$ be its \orderk{k} de~Bruijn graph. Pattern $X \in \Sigma^{\ast}$ matches node $v \in V'$ if it prefix-matches string $G'.\gkey(v)$. The set of occurrences for pattern $X$ is the union of sets $G.\gvalue(v)$ for the matching nodes $v \in V'$.
\end{definition}

As Lemmas~\ref{lemma:dbg-fn} and \ref{lemma:dbg-fp} suggest, there will be no false negatives, and no \emph{false positives} (path labels that exist in the de~Bruijn graph but not in the input graph) with patterns of length at most $k$. With longer patterns, we have to \emph{verify} the results in graph $G$ to avoid false positives. We will develop the idea of using graphs similar to de~Bruijn graphs as path indexes further in the next section.

@@

\begin{definition}[Path graph]
Let $\mathcal{S}$ be a \kcollection{k} of the labels of maximal paths in graph $G = (V, E)$, and let $\mathcal{K}$ be a prefix-free set of strings containing $\#^{k}$ and $\$^{k}$. We assume that each right-maximal substring $S$ in $\mathcal{S}$ prefix-matches a string $K \in \mathcal{K}$, and that each string $K \in \mathcal{K}$ is a substring of $\mathcal{S}$. Then the \orderk{k} \emph{path graph} of graph $G$ with \emph{key set} $\mathcal{K}$ is a graph $G' = (V', E')$ such that
\begin{enumerate}
\item each node $v_{K} \in V'$ represent a distinct key $K \in \mathcal{K}$, with $G'.\glabel(v_{K}) = K[0]$;
\item each node has a key $G'.\gkey(v_{K}) = K$; and
\item each edge $(v_{K}, v_{K'}) \in E'$ represents the occurrence of substring $K[0] K'$ in $\mathcal{S}$ such that strings $K$ and $K[0] K'$ prefix-match.
\end{enumerate}
We use the node corresponding to $\#^{k}$ as the source node $s$ and the node corresponding to $\$^{k}$ as the sink node $t$, adding the technical edge $(t, s)$ in the usual way.
\end{definition}

% FIXME from here
- using path graphs as indexes
- k-index: has a node matching any k-path in the input graph
- edge condition: Y has predecessor with c -> edge (X, Y) if cY matches X
- k-equivalent: union of values for nodes matching X identical for all |X| <= k
- k-equivalent + edge condition: no false positives of length <= k; no false negatives

pruned de Bruijn graphs: merge nodes of DBG while keeping the path graph k-equivalent
similar to manifold de Bruijn graphs \cite{Lin2014}

@@

I thank Richard Durbin, Erik Garrison, and Adam Novak for \dots
